A Bayesian method for latitudinal temperature gradients in the early Eocene

### 2023-02-08 manuscript/manuscript.Rmd
discussion update

# Discussion
The Bayesian model presented herein accurately reconstructs the modern, latitudinal SST gradient from a patchy and relatively small (n = 34) sample of temperature data. This is an advancement over previously used linear or quadratic approximations [e.g. @Refs], which are inherently unable to capture the sigmoidal shape of the modern SST gradient. Our generalised logistic function is flexible enough to approximate the shapes of all of those functions. As such, our model presents an alternative to non-parametric methods for inferring latitudinal temperature gradients. Non-parametric models can very flexibly follow the shape of an unknown temperature gradient [e.g. @Zhang2019; @Jones2022], but are unsuitable for extrapolation, and have no model parameters that can be readily compared between data sets. When used for interpolation, non-paremetric methods tend to strictly respond to the data, even if the resulting patterns are unlike those predicted by physical processes []. In contrast, the four parameters that describe the shape of our generalised logistic model clearly relate to the magnitude, steepness and the latitude of greatest steepness of the modeled gradient. Given that at least vague prior information on the shape of latitudinal temperature gradients on Earth should exist for any geological time period, we argue that parametric methods should be preferred for reconstructing those gradients.

The ease with which such prior information can be integrated is a major advantage of our method: The Bayesian framework allows to specify arbitrary priors on the four parameters that control the shape of the gradient (see Methods). In practice, those priors can be informed by physical constraints, e.g. restricting water temperatures to above the freezing point. A further improvement to existing methods is that the hierarchical structure of our model explicitly accounts for variation of temperature estimates at individual localities, and for uncertainty associated with the ecological temperature proxies.

**Methodological advances.** Latitudinal temperature gradients have been estimated with a variety of parametric and non-parametric statistical techniques. Parametric methods include polynomials [@Greenwood1995; @Bijl2009] and a cosine function [@Inglis2020]. Whilst some of those functions are flexible enough to model a range of empirical temperature gradients, the logistic function used herein has the added advantage of an easily understandable relationship between each of the model parameters and the shape of the latitudinal gradient (Fig. 2). Non-parametric models can very flexibly follow the shape of an unknown temperature gradient [@Jones2022, but see @Zhang2019], but have the disadvantage of having no model parameters that can be compared between data sets. Parametric methods should therefore be preferred to non-parametric temperature gradient models, as long as very broad prior information on the shape of the gradients exists. The ease with which such prior information can be integrated is perhaps the major advantage of our method: The Bayesian framework allows to specify arbitrary priors on the four parameters that control the upper and lower temperature limits, the steepness of the gradient, and the latitudinal position of the steepest point in the gradient (see Methods). A further improvement to existing methods is that the hierarchical structure of our model explicitly accounts for variation of temperature estimates at individual localities, and for uncertainty associated with the ecological temperature proxies. An extension of the model additionally accounts for measurement uncertainty of the geochemical proxies (Supplementary Materials).

-GP regression and interpolation can give rise to inverse gradients ect. --> our method nicely prevents this.

>>>>

# Discussion
The Bayesian model presented herein accurately reconstructs the modern, latitudinal SST gradient from a patchy and relatively small (n = 34) sample of temperature data. This is an advancement over previously used linear or quadratic approximations [e.g. @Refs], which are inherently unable to capture the sigmoidal shape of the modern SST gradient. Our generalised logistic function is flexible enough to approximate the shapes of all of those functions. As such, our model presents an alternative to non-parametric methods for inferring latitudinal temperature gradients, which are sometimes favoured as they can very flexibly follow the shape of an unknown temperature gradient [e.g. @Zhang2019; @Jones2022]. However, when used for interpolation or prediction outside the proxy range, non-parametric methods such as Gaussian process regression strictly respond to the data [e.g. @Inglis2020], and the resulting large-scale patterns are informed by the idiosyncracies of a patchy proxy record.

In parametric models, placing informative priors on parameters can improve interpolation and prediction. For example, the greater amount of solar radiation per unit area in low latitudes causes Earth's latitudinal temperature gradient to be broadly negative [@Beer2008]. Prior information on the shape of latitudinal temperature gradients on Earth thus exists for all geological time periods, and should be used in any palaeoclimate reconstruction. The ease with which such prior information can be integrated is a major advantage of our method: The Bayesian framework allows to specify arbitrary priors on the four parameters that control the shape of the gradient (see Methods). In practice, those priors are informed by physical constraints, e.g. restricting water temperatures to above the freezing point. A further improvement to existing methods is that the hierarchical structure of our model explicitly accounts for variation of temperature estimates at individual localities, and for uncertainty associated with the ecological temperature proxies.

Modelling the latitudinal temperature gradient from the equator to the poles facilitates reconstruction of the global mean temperature.

Our parametric model has four parameters that describe the shape of the modeled gradient, and they clearly relate to its magnitude, steepness and the latitude of its greatest steepness.

**Methodological advances.** Latitudinal temperature gradients have been estimated with a variety of parametric and non-parametric statistical techniques. Parametric methods include polynomials [@Greenwood1995; @Bijl2009] and a cosine function [@Inglis2020]. Whilst some of those functions are flexible enough to model a range of empirical temperature gradients, the logistic function used herein has the added advantage of an easily understandable relationship between each of the model parameters and the shape of the latitudinal gradient (Fig. 2). Non-parametric models can very flexibly follow the shape of an unknown temperature gradient [@Jones2022, but see @Zhang2019], but have the disadvantage of having no model parameters that can be compared between data sets. Parametric methods should therefore be preferred to non-parametric temperature gradient models, as long as very broad prior information on the shape of the gradients exists. The ease with which such prior information can be integrated is perhaps the major advantage of our method: The Bayesian framework allows to specify arbitrary priors on the four parameters that control the upper and lower temperature limits, the steepness of the gradient, and the latitudinal position of the steepest point in the gradient (see Methods). A further improvement to existing methods is that the hierarchical structure of our model explicitly accounts for variation of temperature estimates at individual localities, and for uncertainty associated with the ecological temperature proxies. An extension of the model additionally accounts for measurement uncertainty of the geochemical proxies (Supplementary Materials).
 Disadvanand have no model parameters that can be readily compared between data sets. 
-GP regression and interpolation can give rise to inverse gradients ect. --> our method nicely prevents this.
